
# Patterns Over Time

## Surface Water Temperature

For a set of lakes around the world, lake surface water temperature (LSWT) has been obtained using reflectance measurements recorded by the AATSR satellite system. We have a set of monthly LSWT values for Lake Balaton, a large freshwater lake located in Hungary. Note that the observed series has been reconstructed so that it is possible for LSWT below zero to be recorded.  

<br>

Look at the first few lines of the data frame by running the following code:

```{r headb, eval=FALSE}

head(balaton)

```

<br>

This worksheet contains 4 columns, described as follows: 

* `C1` Year 
* `C2` Month (1= Jan, ..., 12= Dec) 
* `C3` Season (1= Mar-May, 2= Jun-Aug, 3= Sept-Nov, 4= Dec-Feb) 
* `C4` Lake surface water temperature (degrees Celsius) 

<br>

Column names can be seen by running the following line of code:

```{r namesb, eval=FALSE}

names(balaton)

```

<br>

The whole dataset can be viewed by typing the name of the object in the script file and running that line:

```{r b, eval=FALSE}

balaton

```

<br>

## Naming Columns

<br>

We are going to extract and name columns from the data set so they can be easily referred to later on. Run the following lines of code: 

```{r extrct, eval=FALSE}

year   <- balaton[,1] 
month  <- balaton[,2] 
lswt   <- balaton[,4]

year  # this allows you to check what the contents of the object created 

```

<br>

## Decimal Date

<br>

Decimal date provides a continuous measure of time which is useful for producing plots. To construct decimal date we need to create a new variable, say `dec.date`, as per the following formula:

$$\mathrm{dec.date} = \mathrm{year}+\frac{\mathrm{month}-1}{12} $$
<br>

### **Task 1**


**Write code to create a new variable called `dec.date` from the variables `year` and `month`.**

`r hide("Hint")`

You may want to write the above formula in R.

`r unhide()`

`r hide("Solution")`

```{r sol1, eval=FALSE}

dec.date <- year + (month-1)/12


```

`r unhide()`

You can check your variable by simply running 'dec.date' in your R script, after having created the variable. 

<br>

## Plotting The Data

<br>

For a single lake, the **pattern over time** for LSWT can be examined by plotting the data. (We first need to add the newly created 'dec.date' varaible to the data set). 

```{r timeplot, eval=FALSE}

rename(balaton, lswt = BALATON) # Renaming the variable in the data set from it's original BALATON to lswt

balaton <- balaton %>% 
  mutate(dec.date = year + (month-1)/12) # Adding the dec.date variable 


ggplot(data = balaton, aes(x = dec.date, y = lswt)) + 
  geom_line() +
  labs(title = "LSWT (Celsius) at Lake Balaton") +
  xlab("Date") +
  ylab("LSWT (Celcius)")

# Creating a line plot using the tidyverse packages and our variables
```

<br>

**Which of the following sentences best describes the LSWT patterns over time at Lake Balaton?**

```{r, echo = FALSE}
lswt_pat <- sample(c(
    "Evidence for both trend and seasonal pattern",
    answer = "Evidence for a seasonal pattern but no evidence for a trend",
    "No evidence for regular seasonal pattern",
    "Evidence for an increasing trend")
)
```

`r longmcq(lswt_pat)`

<br>
