
# Seasonal Patterns

In addition to looking at trend over time it is also of interest to examine the seasonal pattern within each year. 

## **Task 2**


**Write code to produce a seasonal plot of the LSWT.**

`r hide("Hint")`

You may want to copy the previous code for exploring the patterns over time but use the `month` variable instead, and change any other necessary code. You will also want to produce a scatter plot as opposed to a line graph.
`r unhide()`

`r hide("Solution")`

```{r sol2, eval=FALSE}

ggplot(data = balaton, aes(x = month, y = lswt)) + 
  geom_point() +
  labs(title = "Monthly LSWT (Celsius) at Lake Balaton") +
  xlab("Month") +
  ylab("LSWT (Celcius)")

```

`r unhide()`

<br>

## **Task 3**


**Write code to produce a boxplot of the LSWT by month**

`r hide("Hint")`

You may want to use 'geom_boxplot' instead of 'geom_line'.

`r unhide()`

`r hide("Solution")`

```{r sol3, eval=FALSE}

ggplot(data = balaton, aes(x = as.factor(month), y = lswt)) + 
  geom_boxplot() +
  labs(title = "Monthly LSWT (Celsius) at Lake Balaton") +
  xlab("Month") +
  ylab("LSWT (Celcius)")

```

`r unhide()`

<br>

Boxplots enable us to see a summary of the LSWT measurements for each month, and they also make it easy to compare the distribution and variability of lswt across the months.

<br>

**In which months are the lowest temperatures observed?**

```{r, echo = FALSE}
lowq <- sample(c(
    "November - December",
    "July - August",
    answer = "January - February",
    "December - January"
   )
)
```

`r longmcq(lowq)`

<br>
